---
title: "P8105 Homework 2"
author: "Astrid Xu"
date: "2026-10-06"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

# Question 1
```{r}
subway =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11,
    entry, vending, entrance_type, ada)
subway

subway =
  subway |>
  mutate(
    entry = ifelse(entry == "YES", TRUE, FALSE))
subway

# Could also use case_match
# mutate(
#  entry = case_match(
#    entry,
#    "YES" ~ TRUE,
#    "NO" ~ FALSE
# ))
```
After cleaning, there are 1868 observations and 19 variables in the dataset. The dataset contains information about subway lines, station names, station locations (latitude and longitude), routes served, whether entrances allow entry, vending availability, entrance types, and ADA compliance. For the data cleaning steps, I first imported the dataset using read_csv() and cleaned all variable names using clean_names(). The original dataset contained 32 variables in total, so I used select() to retain only the required variables and saved the cleaned dataset as subway. I then checked the dataset to make sure that it satisfied the requirements. Next I used mutate() to modify the existing entry variable. I am more familiar with ifelse(), it means if the observation equals to YES, then convert it to TURE. If else other than YES, then convert it to FALSE. I viewed the subway dataset again to check the result. The resulting dataset is not tidy because route information is stored across multiple variables (route1 through route11) instead of having route number and route name as distinct variables.

```{r}
subway |>
  distinct(line, station_name)
```
There are 465 distinct stations.

```{r}
subway |>
  filter(ada == TRUE) |>
  count(line, station_name)
```
Only 84 stations are ADA compliant.

```{r}
no_vending_entry =
  subway |>
  filter(vending == "NO", entry == TRUE)

no_vending =
  subway |>
  filter(vending == "NO")

nrow(no_vending_entry) / nrow(no_vending)
```
The proportion of station entrances / exits without vending allow entrance is 0.3770.

```{r}
# Reformat can't be proceeded, cuz the variable type of routes doesn't match. So fix the variable types first
str(subway)
subway =
  subway |>
  mutate(
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
  )
str(subway)

subway_reform =
  subway |>
  pivot_longer(route1:route11,
    names_to = "route_number",
    values_to = "route_name")
subway_reform

A_train =
  subway_reform |>
  filter(route_name == "A") |>
  distinct(line, station_name)
A_train

A_train_ada =
  subway_reform |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name)
A_train_ada
```
60 distinct stations serve the A train. Of those stations that serve the A train, only 17 are ADA compliant.


# Question 2
```{r}
trash =
  read_excel("data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742"
  )
trash

trash =
  trash |>
  janitor::clean_names()
names(trash)

trash =
  trash |>
  mutate(
    sports_balls = round(sports_balls),
    sports_balls = as.integer(sports_balls)
  )
trash

str(trash)
```

```{r}
professor =
  read_excel("data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M139"
  )
professor

professor =
  professor |>
  janitor::clean_names()
names(professor)

str(professor)
```

```{r}
gwynnda =
  read_excel("data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L394"
  )
gwynnda

gwynnda =
  gwynnda |>
  janitor::clean_names()
names(gwynnda)

str(gwynnda)
```

```{r}
trash =
  trash |>
  mutate(
    trash_wheel = "Mr. Trash Wheel"
  )

professor =
  professor |>
  mutate(
    trash_wheel = "Professor Trash Wheel"
  )

gwynnda =
  gwynnda |>
  mutate(
    trash_wheel = "Gwynnda"
  )

# Combination can't be proceeded, cuz year and plastic_bottles have two different variable types (chr&dbl), so need to fix them first
trash =
  trash |>
  mutate(
    year = as.numeric(year)
  )

professor |>
  filter(plastic_bottles == "dive")

professor =
  professor |>
  mutate(
    plastic_bottles = ifelse(plastic_bottles == "dive", NA, plastic_bottles),
    plastic_bottles = as.numeric(plastic_bottles)
  )
professor

trash_combined =
  bind_rows(trash, professor, gwynnda)
trash_combined
```
The merged dataset contains 1,269 observations and 15 variables. It tells key information including the name of the garbage truck, the trash can number, the month, year, and date of collection, as well as the weight and volume of the collected trash. The dataset also provides the specific types of trash collected, such as plastic bottles, cigarette butts, plastic bags, and wrapping paper.

```{r}
professor_total =
  trash_combined |>
  filter(trash_wheel == "Professor Trash Wheel")

professor_total |>
  summarize(
    total_weight = sum(weight_tons)
  )
```
The total weight of trash collected by Professor Trash Wheel was 295 tons.

```{r}
gwynnda_june_total =
  trash_combined |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  )

gwynnda_june_total |>
  summarize(
    total_cigarette_butts = sum(cigarette_butts)
  )
```
The total number of cigarette butts collected by Gwynnda in June 2022 was 18,120.

# Question 3
```{r}
baseline =
  read_csv(
    "data/MCI_baseline.csv",
    skip = 1
  )
baseline

baseline =
  baseline |>
  janitor::clean_names()
names(baseline)

baseline =
  baseline |>
  mutate(
    sex = case_match(
      sex,
      1 ~ "Male",
      0 ~ "Female"
    ),
    apoe4 = case_match(
      apoe4,
      1 ~ "Carrier",
      0 ~ "Non-carrier"
    )
  )
baseline

baseline =
  baseline |>
  mutate(
    age_at_onset = na_if(age_at_onset, "."),
    age_at_onset = as.numeric(age_at_onset)
  )
baseline
```
After importing the dataset using read_csv(), I noticed that the first row of the table did not contain the variable names I needed, but instead contained descriptions of the variables. So to make R skip the first row and recognize the second row as the variable names, I added the skip = 1 argument when using read_csv(). Then based on the codebook, I converted sex and APOE4 carrier status into their corresponding categorical variables and converted missing values to NA. To remove participants who did not meet the stated inclusion criteria (i.e., no MCI at baseline), I first used filter() to identify participants who had already developed MCI by the time they entered the study. After checking these participants individually, I used filter() to retain only participants whose current age was less than their age at MCI onset or who did not develop MCI during the follow-up period. The original dataset has 483 participants, resulting dataset contains 479 participants and 6 variables, including participant ID, baseline age, sex, education, APOE4 carrier status, and age at MCI onset.

```{r}
baseline |>
  filter(age_at_onset <= current_age)

baseline =
  baseline |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)
baseline

developed =
  baseline |>
  filter(!is.na(age_at_onset))
nrow(developed)
```
After excluding, 479 participants were recruited into the study. Among them, 93 developed MCI during the follow-up period.

```{r}
baseline |>
  summarize(
    average_baseline_age = mean(current_age)
  )
```
The average baseline age was 65.0 years.

```{r}
female =
  baseline |>
  filter(sex == "Female")

female_carrier =
  baseline |>
  filter(
    sex == "Female",
    apoe4 == "Carrier"
  )

nrow(female_carrier) / nrow(female)
```
Among female participants, 30% were APOE4 carriers.

```{r}
amyloid =
  read_csv(
    "data/mci_amyloid.csv",
    skip = 1
  )
amyloid

amyloid =
  amyloid |>
  janitor::clean_names()
names(amyloid)

str(amyloid)

amyloid |>
  filter(is.na(as.numeric(baseline)))

amyloid =
  amyloid |>
  mutate(
    baseline = as.numeric(baseline),
    time_2 = as.numeric(time_2),
    time_4 = as.numeric(time_4),
    time_6 = as.numeric(time_6),
    time_8 = as.numeric(time_8)
  )
amyloid

str(amyloid)

amyloid =
  amyloid |>
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    values_to = "amyloid"
  )
amyloid
```
After importing the data, I noticed that the rows and variable names had the same issue as the baseline dataset. So I again used skip = 1 in read_csv() to skip the first row and use the second row as the variable names. After cleaning the variable names, I noticed that the biomarker values which should be numeric were recognized as character variables. I suspected that this was caused by some non-numeric values in these columns. I used is.na() with numeric conversion to identify values that could not be converted to numbers and found that missing values were recorded in two different ways, "NA" and "Na". Since both represented missing values, I directly converted all biomarker variables to numeric values, which converted these values to NA without affecting the available measurements. Then I used pivot_longer() to separate the five measurement times for each participant into individual observations. The original dataset contained 487 participants and 6 variables, with five amyloid biomarker measurements stored in separate columns for baseline and Times 2, 4, 6, and 8. After reformatting, the dataset contains 2,435 observations and 3 variables: participant ID, measurement time, and amyloid biomarker value. Each row now represents one participant at one measurement time.

```{r}
amyloid =
  amyloid |>
  rename(id = study_id)

mci =
  baseline |>
  inner_join(amyloid, by = "id")
mci

mci |>
  distinct(id) |>
  nrow()
```
I used an inner join to merge the two datasets by participant ID. This matches observations with the same participant ID across the two datasets, while participants who are not present in both datasets are excluded. So the resulting dataset contains only participants who have information in both the baseline and amyloid datasets. The final merged dataset contains 2,355 observations and 8 variables, representing 471 unique participants. Each participant has five observations corresponding to the five amyloid measurement times.

```{r}
write_csv(
  mci,
  "data/mci_merged.csv"
)
```

